Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-03-17 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update popularity case study with ggmice functions

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Mu\~noz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article: default
    word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph, linking multilevel equations to \pkg{lme4} formulas.]

@@

@@

# How (not) to impute (Case Study I: Popularity)

@@

TODO: make a 'translation table' between model equation and lme4. Use betas instead of gamma's and mu's.

@@

```{r}
popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
```

```{r pop_pat, echo=FALSE, fig.cap = "Missing data pattern in the popularity data", message=FALSE, warning=FALSE}
# missingness
plot_md_pat(popmis)
```

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`.

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight only one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend (pop obs and pop miss)
# TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon
# TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?)

@@

Complete case analysis ignores the observations with missingness altogether, which may even introduce bias in MCAR situations.

@@

The first imputation model that we'll use is likely to be invalid. In this model, we ignore the multilevel structure of the data, despite the high ICCs. And assumes exchangeability between units. This is purely to illustrate the effects of ignoring the clustering in our imputation effort.

We'll use predictive mean matching to impute the continuous variables and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `school` as predictors to impute other variables.

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# dry run to get imputation parameters
ini <- mice(popmis, maxit = 0)

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("school")] <- 0

@@

As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `school`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the school.

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred <- ini$pred 
# pred[, "pupil"] <- 0

# impute the data, cluster as predictor
imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)

@@

```{r pop_norm}
pred <- ini$pred
pred["popular", ] <- c(-2, 2, 2, 2) 
#-2 for the cluster variable, 2 for random effects
meth <- ini$meth
meth["popular"] <- "2l.pmm"
imp_pmm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: "Johanna Mu\\~{n}oz Avila"
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  short:     "Multilevel \\pkg{mice}"
abstract: >
  This is a tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Footnotes in the current version show work in progress/under construction. The last section is not part of the manuscript, but purely for reminders. We aim to submit at JSS, so there is no word count limit ("There is no page limit, nor a limit on the number of figures or tables").
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
header-includes: 
 - \usepackage{graphicx} 
 - \usepackage{mathtools}
 - \usepackage{ulem}
output: 
    rticles::jss_article:
      keep_tex: yes
      number_sections: yes
    # word_document: default
bibliography: ../References/multilevelmice.bib
editor_options: 
  chunk_output_type: inline
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
options(prompt = 'R> ', continue = '+ ')

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas. Use betas instead of gamma's and mu's. Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).]

@@

Set-up the R environment and load the necessary packages:

```{r}
set.seed(2022)
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
library(lme4)
```

@@

# Case Study I: How (not) to impute

@@

@@

```{r}
popmis <- popmis[, c("school", "popular", "sex", "texp")] 
```
Plot the missing data pattern:

```{r pop_pat, fig.cap = "Missing data pattern in the popularity data"}
plot_pattern(popmis)
```

The missingness is univariate and sporadic, which is illustrated in the missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the incomplete data is `r round(icc(popular ~ as.factor(school), data = na.omit(popmis)), 2)`. This tells us that the multilevel structure of the data should probably be taken into account. If we don't, we'll may end up with incorrect imputations, biasing the effect of the clusters towards zero.

Plot the correlations in the incomplete data:

```{r pop-corr}
plot_corr(popmis)
```

To develop the best imputation model for the incomplete variable `popular`, we need to know whether the missingness depends on the observed values of other variables. We'll highlight one other variable to illustrate, but ideally one would inspect all relations. The questions we'll ask are: 'Does the missing data of pupil popularity (`popular`) depend on observed teacher popularity (`texp`)?'. This can be evaluated statistically, but visual inspection usually suffices. We'll make a histogram of `texp` separately for the pupils with known popularity and missing popularity.

Plot the histogram for teacher experience conditional on the missingness indicator of `popular`:

```{r pop-hist}
ggmice(popmis, aes(texp)) +
  geom_histogram(fill = "white") +
  facet_grid(. ~ is.na(popular), labeller = label_both)
```

```{r pop-hist-clust, include=FALSE}
# popmis %>% 
#   filter(school < 5) %>% 
# ggmice(., aes(texp)) +
#   geom_histogram(fill = "white") +
#   facet_grid(school ~ is.na(popular), labeller = label_both)
# # does not make sense because texp is a cluster level variable!
```

This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` ($t = $ `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).

<!-- In Figure \ref{fig:pop_dist} we see that **[update this part]** the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. (In fact, this is exactly what happens, because the missingness in these data was created manually.) We've made it observable by examining the relations between the missingness in popular and the observed data in `texp`.  -->

```{r pop_dist, fig.cap = "Conditional distributions in the popularity data", echo=FALSE, message=FALSE, warning=FALSE}
# plot_conditional(popmis, x = "texp", z = "popular", cluster = "school")
# TODO: check if up-side-down plot works
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
# TODO: add title and informative legend (pop obs and pop miss)
# TODO: add facets for some clusters, or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal, alternatively add a quartile line around the density with geom_ribbon
# TODO: add fill and/or some way of expressing how much the densities overlap (overlap coefficient in %?)

@@

Complete case analysis ignores the observations with missingness altogether, which lowers statistical power and may even introduce bias in MCAR situations.

@@

The first imputation model that we'll use is likely to be invalid. We do *not* use the cluster identifier `school` as imputation model predictor. With this model, we ignore the multilevel structure of the data, despite the high ICC. This assumes exchangeability between units. We include it purely to illustrate the effects of ignoring the clustering in our imputation effort. We'll use the default imputation methods in `mice()` (predictive mean matching to impute the continuous variables and logistic regression to impute binary variables).

Create a predictor matrix and remove `school` as predictor for `popular`:

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# predictor matrix
pred <- quickpred(popmis)
plot_pred(pred)

@@

<!-- We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each school has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the school.  -->

@@

```{r pop_predictor, message=FALSE, warning=FALSE}
# adjust the predictor matrix
pred["popular", "school"] <- 1 
plot_pred(pred)

# impute the data, cluster as predictor
imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

@@

```{r pop_norm}
# adjust the predictor matrix
pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
plot_pred(pred) 
meth <- make.method(popmis)
meth["popular"] <- "2l.pmm"
imp_pmm_2l <-
  mice(
    popmis %>% mutate(school = as.integer(school)),
    pred = pred,
    meth = meth,
    maxit = 1,
    print = FALSE
  )
```

@@

- Show use case for deductive imputation for cluster level variables?

### 2022-03-17 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update popularity case study with ggmice functions

@@

\author{
Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde noz
Avila\\University Medical Center Utrecht \AND Valentijn de
Jong\\University Medical Center Utrecht \And Gerko Vink\\Utrecht
University \AND Thomas Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde noz
Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph,
  linking multilevel equations to \pkg{lme4} formulas.}

@@

\hypertarget{how-not-to-impute-case-study-i-popularity}{%
\section{How (not) to impute (Case Study I:
Popularity)}\label{how-not-to-impute-case-study-i-popularity}}

@@

TODO: make a `translation table' between model equation and lme4. Use
betas instead of gamma's and mu's.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> popmis <- mice::popmis[, c("school", "popular", "sex", "texp")] 
\end{CodeInput}
\end{CodeChunk}

@@

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
incomplete data is 0.56.

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight only one other
variable to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{texp})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{texp}
separately for the pupils with known popularity and missing popularity.

In Figure \ref{fig:pop_dist} we see that \textbf{{[}update this part{]}}
the distribution for the missing \texttt{popular} is further to the
right than the distribution for observed \texttt{popular}. This would
indicate a right-tailed MAR missingness. (In fact, this is exactly what
happens, because the missingness in these data was created manually.)
We've made it observable by examining the relations between the
missingness in popular and the observed data in \texttt{texp}.

{\centering \includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_dist-1}

\caption[Conditional distributions in the popularity data]{Conditional distributions in the popularity data}\label{fig:pop_dist}
\end{figure}
\end{CodeChunk}

@@

The first imputation model that we'll use is likely to be invalid. In
this model, we ignore the multilevel structure of the data, despite the
high ICCs. And assumes exchangeability between units. This is purely to
illustrate the effects of ignoring the clustering in our imputation
effort.

We'll use predictive mean matching to impute the continuous variables
and logistic regression to impute the binary variable \texttt{sex}. We
do not use the observation identifier \texttt{pupil} or cluster
identifier \texttt{school} as predictors to impute other variables.

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(popmis, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("school")] <- 0
R> 
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

As the original ICCs show, 100\% of the variance in \texttt{texp} can be
attributed to the clustering variable \texttt{school}. This tells us
that the multilevel structure of the data should be taken into account.
If we don't, we'll end up with incorrect imputations, biasing the effect
of the clusters towards zero.

We can also observe that the teacher experience increases slightly after
imputation. This is due to the MNAR missingness in \texttt{texp}. Higher
values for \texttt{texp} have a larger probability to be missing. This
may not a problem, however, if at least one pupil in each school has
teacher experience recorded, we can deductively impute the correct
(i.e.~true) value for every pupil in the school.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> # pred[, "pupil"] <- 0
R> 
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- ini$pred
R> pred["popular", ] <- c(-2, 2, 2, 2) 
R> #-2 for the cluster variable, 2 for random effects
R> meth <- ini$meth
R> meth["popular"] <- "2l.pmm"
R> imp_pmm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  Adding some kind of help function to mice that suggests a suitable
  predictor matrix to the user, given a certain analysis model.
\item
  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
\item
  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\item
  Adding a ICC=0 dataset to show that even if there is no clustering it
  doesn't hurt.
\item
  env dump in repo
\end{itemize}

>>>>

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

@@

\author{
Hanne Oberman\\Utrecht University \And Johanna Mu\textasciitilde\{n\}oz
Avila\\University Medical Center Utrecht \AND Valentijn de
Jong\\University Medical Center Utrecht \And Gerko Vink\\Utrecht
University \AND Thomas Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Mu\textasciitilde\{n\}oz
Avila, Valentijn de Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph
  or `translation table' linking multilevel equations to \pkg{lme4}
  formulas. Use betas instead of gamma's and mu's. Add interpretation of
  values in predictormatrix (-2 for the cluster variable, 2 for random
  effects).}

Set-up the R environment and load the necessary packages:

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)
R> library(ggmice)
R> library(ggplot2)
R> library(dplyr)
R> library(lme4)
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{case-study-i-how-not-to-impute}{%
\section{Case Study I: How (not) to
impute}\label{case-study-i-how-not-to-impute}}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> popmis <- popmis[, c("school", "popular", "sex", "texp")] 
\end{CodeInput}
\end{CodeChunk}

Plot the missing data pattern:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(popmis)
\end{CodeInput}
\begin{figure}

@@

The missingness is univariate and sporadic, which is illustrated in the
missing data pattern in Figure \ref{fig:pop_pat}. The ICC in the
incomplete data is 0.56. This tells us that the multilevel structure of
the data should probably be taken into account. If we don't, we'll may
end up with incorrect imputations, biasing the effect of the clusters
towards zero.

Plot the correlations in the incomplete data:

\begin{CodeChunk}
\begin{CodeInput}
R> plot_corr(popmis)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-corr-1} \end{center}

To develop the best imputation model for the incomplete variable
\texttt{popular}, we need to know whether the missingness depends on the
observed values of other variables. We'll highlight one other variable
to illustrate, but ideally one would inspect all relations. The
questions we'll ask are: `Does the missing data of pupil popularity
(\texttt{popular}) depend on observed teacher popularity
(\texttt{texp})?'. This can be evaluated statistically, but visual
inspection usually suffices. We'll make a histogram of \texttt{texp}
separately for the pupils with known popularity and missing popularity.

Plot the histogram for teacher experience conditional on the missingness
indicator of \texttt{popular}:

\begin{CodeChunk}
\begin{CodeInput}
R> ggmice(popmis, aes(texp)) +
+   geom_histogram(fill = "white") +
+   facet_grid(. ~ is.na(popular), labeller = label_both)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-hist-1} \end{center}

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (\$t = \$ -0.873, p = 0.383).

@@

The first imputation model that we'll use is likely to be invalid. We do
\emph{not} use the cluster identifier \texttt{school} as imputation
model predictor. With this model, we ignore the multilevel structure of
the data, despite the high ICC. This assumes exchangeability between
units. We include it purely to illustrate the effects of ignoring the
clustering in our imputation effort. We'll use the default imputation
methods in \texttt{mice()} (predictive mean matching to impute the
continuous variables and logistic regression to impute binary
variables).

Create a predictor matrix and remove \texttt{school} as predictor for
\texttt{popular}:

\begin{CodeChunk}
\begin{CodeInput}
R> # predictor matrix
R> pred <- quickpred(popmis)
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored-1} \end{center}

\begin{CodeInput}
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred["popular", "school"] <- 1 
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_predictor-1} \end{center}

\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred["popular", ] <- c(school = -2, popular = 0, sex = 2, texp = 2) 
R> plot_pred(pred) 
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_norm-1} \end{center}

\begin{CodeInput}
R> meth <- make.method(popmis)
R> meth["popular"] <- "2l.pmm"
R> imp_pmm_2l <-
+   mice(
+     popmis %>% mutate(school = as.integer(school)),
+     pred = pred,
+     meth = meth,
+     maxit = 1,
+     print = FALSE
+   )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  Adding some kind of help function to mice that suggests a suitable
  predictor matrix to the user, given a certain analysis model.
\item
  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
\item
  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\item
  Adding a ICC=0 dataset to show that even if there is no clustering it
  doesn't hurt.
\item
  Show use case for deductive imputation for cluster level variables?
\item
  env dump in repo
\end{itemize}
